Modelling BCG vaccination in the UK

### 2019-08-05 rmds/98-discussion.Rmd
thesis discussion updated

This thesis has assessed the impact of the 2005 change in BCG vaccination policy in some detail. The aim of this chapter is to provide an overview of the principle findings of this thesis; interpret these findings; discuss the overall strengths and weaknesses of this thesis; outline the potential implications; explore the opportunities for public engagement that this work allowed; and describe potential future research. Each results chapter contains a detailed discussion of the approach used, and the results, for that chapter. Consequently, the aim of this chapter is to summarise and discuss the findings from this thesis as a whole.

@@

In Chapter \@ref(data), I found that there was some evidence that negative outcomes were more frequent in Tuberculosis (TB) cases not BCG vaccinated than in those that were. I also found that missingness in routine surveillance sources of TB data was associated with multiple risk factors. In Chapter \@ref(beneficial-bcg-out), I found supporting evidence that BCG vaccination was associated with reduced all-cause morality with some evidence that this may have been due to reduced TB mortality. I found little evidence for any other association with TB outcomes, after adjusting for confounding. In Chapter \@ref(sutherland), I explored some of the evidence that was used by policy makers to assess the impact of the change in policy. I found that the previous approach had drastically underestimated the amount of uncertainty surrounding the effect estimates. Using newly available data, I also found that ending the scheme was projected to result in greater number of notifications in the UK born than previously thought. These findings were confirmed in Chapter \@ref(direct-eff), where I evaluated the evidence in the surveillance data that the change in policy had impacted TB incidence rates in the target populations. However, in this chapter I found that any increase in TB notifications in the UK born was likely far outweighed by reductions in the number of notifications in the non-UK born. Although I was unable to rule out an unrelated policy change. Finally in Chapter \@ref(model-results), I found that the BCG schools' scheme was projected to reduce overall TB incidence compared to both neonatal and no vaccination over a range of time horizons. However, neonatal vaccination reduced incidence in children compared to any other scenario although it had little impact in any other age group. No vaccination programme evaluated had an impact on incidence in older adults. The results from Chapter \@ref(model-results) are preliminary only as the model on which they were based was a very poor fit to the observed data (Chapter \@ref(model-fitting)).

@@

This thesis has used multiple methods, and data sources, to explore the impact of the 2005 change in BCG policy. This triangulation allows for more certainty in the findings than if only a single approach had been used. A limitation of the work in this thesis is that all results were based on surveillance data. Surveillance data is subject to multiple bias issues (see Chapter \@ref(data) for details). Ideally multiple different data types would have been used to more effectively triangulate the impact of the change in policy.  However, the data used in this thesis represents the best available. No similarly thorough use of an equivalent data source exists for TB. Another limitation of the work in this thesis is the very poor fit of the dynamic model, developed in Chapter \@ref(model-development), to the observed data. This made drawing conclusions from it difficult. On the other hand, the model fitting presented in Chapter \@ref(model-fitting) represented one of the only examples of fitting a very complex model to data using robust bayesian, plug and play, approaches. The lack of success is itself a useful result from which much can be learned about the usability of the fitting tools and the upper limits on model complexity. A major strength of the work in this thesis is the attention that has been paid to make it both open and reproducible. Hopefully, this will allow these findings to be more easily validated, and built upon, by others. Finally the work in this thesis generated several tools as a by-product of the main research question. These tools have benefited other research projects.

@@

This thesis has highlighted the trade-off between vaccinating those at school-age and neonates in settings where the waning of BCG effectiveness when given later in life is minimal. Whilst policy makers were previously aware of this trade-off, their was little quantitative evidence exploring it explicitly. Globally, BCG policy does not account for areas where the BCG vaccine may be equally effective regardless of when it is given.[@TheWorldHealthOrganization:2018va] The evidence from this thesis should, therefore, be considered when outlining future BCG policy. In addition, new TB vaccines are in development that may be less susceptible to waning effectiveness when given later in life over a greater geographic area.[@Medicine2013] The findings from this thesis may be applicable to these vaccines in areas where the BCG vaccine is currently known to be ineffective. This may mean that these newly developed vaccines may be better targeted at those at school-age, rather than neonates, depending on the duration of protection that they provide. The work from Chapter \@ref(model-development) and Chapter \@ref(model-fitting) may be particularly suitable to adaption for this use case. This thesis has also explored the potential benefits of BCG vaccination on TB outcomes. The evidence of a reduction in all-cause TB mortality in TB cases may add additional weight to the argument that wider vaccination maybe more cost effective than previously thought in low incidence countries. These findings may also be used to drive vaccine uptake as they provide additional incentives for vaccination. Finally, this thesis has shown the that the impact of the BCG policy has varied depending on UK birth status. This may further strengthen the case for varying vaccination policies depending on the country of origin of the target of vaccination policy, and their immediate families country of origin.

@@

Open reproducible research has been a primary focus of this thesis. A version controlled archive of this thesis is available from GitHub^[Thesis GitHub: https://github.com/seabbs/thesis], with a formatted version available on my personal site^[Thesis site: https://www.samabbott.co.uk/thesis]. This thesis relies on data from the Enhanced TB Surveillance system and the Labour Force Survey. The cleaning and munging of this data has been standardised as an R package, `tbinenglanddataclean`^[`tbinenglanddataclean`: https://www.samabbott.co.uk/tbinenglanddataclean/], and is available for download. All chapters that contain analysis are linked to their own GitHub repositories, each of which is fully reproducible (discounting the raw data which cannot be released due confidentiality reasons). Literate coding was used to link analysis code with documentation using the R tool chain. An R package, `prettypublisher`^[`prettypublisher`: https://github.com/seabbs/prettypublisher], was developed to augment these tools. Where possible open source tooling has been used to provide a working analytical environment for each chapter. All chapters that have been peer reviewed, or are undergoing peer review, have been preprinted. The model developed in Chapter \@ref(model-development) has been released as an R package along with the fitting pipeline developed in Chapter \@ref(model-fitting). Tools used to develop the figures in Chapter \@ref(background), using World Health Organization data, were expanded into an R package (see Chapter \@ref(getTBinR)). Tooling developed alongside this thesis follows open source best practices. See Chapter \@ref(introduction) for full details on the open source projects developed as part of this thesis.

@@

Public engagement has been a constant theme throughout my doctoral work. This is closely linked with the previous aim of open and reproducible research. Effort has been taken so that all peer reviewed content is available for the wider public, with Twitter used disseminate findings. Where appropriate, interactive applications have been developed that seek to explore some of the key findings of this thesis, as well as teaching more theoretical concepts used throughout (see Chapter \@ref(introduction)). Numerous case studies have also been produced that outline these theoretical concepts using some of the open source tools developed alongside this thesis^[See my personal site: https://www.samabbott.co.uk]. These tools were themselves developed to lower the barrier of entry to infectious disease research. One of these tools, `idmodelr`, has been released to CRAN. Finally, in 2017 I spent a week at the Green Man Festival exploring the mathematics of vaccination with the general public. This made use of several simple games, as well as an interactive online tool^[Available here: https://github.com/seabbs/pebblegame].

@@

The finding that BCG vaccination may reduce mortality in TB cases from Chapter \@ref(beneficial-bcg-out) require validation in other data sources and settings. A larger sample size may be required in order to unpick the association between BCG vaccination and the cause of mortality. These findings could also be included in a cost effectiveness study of the BCG vaccine. The dynamic model developed in Chapter \@ref(model-development) and fitted in \@ref(model-fitting) did not fit the observed data well. Additional compute time is needed to diagnosis whether this is a limitation of the fitting technique used or if they model itself was a poor fit for the data. Additional strategies for exploring this issue are discussed in Chapter \@ref(model-fitting). In addition this modl does not currently include targeted vaccination of neonates and is only fitted to data up to 2004. In order to be able to more accurately explore current, and future, vaccination policy the extension of the model to the present data would be required. This would potentially be of great use for policy makers. An alternative would be to develop a comparable models in different settings. This would allow the generalisability of the findings to be explored. The dynamic model could also be further generalised to include hypothetical future vaccines with differing characteristics. This would allow vaccine characteristics and optimal deployment strategies to be explored, via simulations, ahead of further development. Both `getTBinR` and `idmodelr` have active user bases and further developments are planned. This includes: additional tooling, documentation, and case studies. Further development of several of the interactive tools discussed in Chapter \@ref(introduction) is also planned.

@@

This thesis has provided new evidence regarding the use of BCG vaccination in England. A simulation study that was used as part of the quantitative evidence for the change in policy was recreated and updated. The results from this updated model suggested that the change in policy was likely to have a greater impact on the UK born, at school-age, than previously thought. This finding was supported by a regression modelling study on the impact on TB incidence rates from the policy change. However, this study also found that the change in policy was associated with some benefits in UK born neonates and a much larger reduction in TB incidence rates in both non-UK born neonates and those at school-age. An additional regression study looking at the possible link between BCG vaccination and improved TB outcomes found some evidence that BCG vaccination was associated with reduced all-cause mortality, with little evidence of any other benefits. This result strengthens the case for wider vaccination. Additionally, a dynamic model of TB transmission was developed to provide a more detailed tool for evaluating the impact of the change in policy. Although this model was a poor fit to the observed data there was some evidence that continuing school-age vaccination would have resulted in fewer overall TB cases but that neonatal vaccination reduced TB incidence in children compared to continuing school-age vaccination. There was also some evidence that none of the vaccination strategies evaluated impacted long term incidence rates in older adults. These findings suggest a stronger case for the use of the BCG vaccine in school-age populations; in areas with an equivalent level of TB transmission to England; and where the effectiveness of the BCG vaccine has been shown to not reduce with age. They also indicate that a future vaccine, without the reduced effectiveness observed in some geographic areas, may be more effectively targeted at those at school-age that at neonates. However, this depends on the potential duration of protection inferred by vaccination. In addition, they highlight the trade-off between neonatal and school-age vaccination with school-age vaccination reducing overall TB incidence but neonatal vaccination reducing incidence in young children where more severe outcomes are more common. The findings from this thesis may be of use to policy-makers to inform vaccine usage both in the UK and globally. As a by-product of the work conducted in this thesis several open source tools have been developed. These tools maybe used as learning resources, for public engangement, and as part of other research projects.

>>>>

This thesis has assessed the impact of the 2005 change in BCG vaccination policy in some detail. The aim of this chapter is to provide an overview of the principle findings of this thesis; interpret these findings; discuss the overall strengths and weaknesses of this thesis; outline the potential implications; explore the opportunities for public engagement that this work allowed; and describe potential future research. Each results chapter contains a detailed discussion of the approach used, the results, and the strengths and weaknesses of the findings, for that chapter. Consequently, the aim of this chapter is to summarise and discuss the findings from this thesis as a whole.

Bacillus Calmette–Guérin (BCG) remains the only licensed vaccine against Tuberculosis (TB). In 2005, England changed from universal vaccination of school-age children to targeted vaccination of high-risk neonates. Little work has been done to assess the impact of this policy change. This thesis evaluates the impact of this change. \par

Whilst the characteristics of TB in England have been reported elsewhere, little attention has been given to the role of BCG. Consequently, I explored and combined, the available data sources. Reporting on data quality issues, trends in incidence rates and differences in outcomes stratified by BCG status. \par

Prior to the change in policy, several studies were carried out to assess the impact. I recreated one such study and found that there was a greater impact than previously thought. \par

Determining the benefits of being BCG vaccinated is necessary to properly assess the impact of the policy change. I evaluated the evidence that vaccination may improve outcomes for TB cases in England and found that there was some evidence of an association between vaccination and reduced mortality. \par

Surveillance data can help assess whether changes in vaccination policy have influenced incidence rates. I used surveillance data to determine whether those at school-age, or neonates, were directly affected by the policy change. I found the policy change was associated with increased notifications in the UK born but this was outweighed by a reduction in notifications in the non-UK born. \par

Statistical modelling is restricted by the available data. Therefore, I developed a dynamic model of TB, fit to available data, to forecast the impact of the policy change. Although the fit to the data was poor the forecasts suggested that continuing school-age vaccination reduced TB incidence compared with neonatal vaccination. Neonatal vaccination reduced incidence in children but had little impact on other age groups.\par

@@

In Chapter \@ref(data), I explored TB epidemiology in England with a focus on BCG vaccination and data completeness. I found that there was some evidence that negative outcomes were more frequent in Tuberculosis (TB) cases not BCG vaccinated than in those that were. I also found that missingness in routine surveillance sources of TB data was associated with multiple risk factors. In Chapter \@ref(beneficial-bcg-out), I used logistic regression to estimate associations between BCG vacination and TB outcomes. I found supporting evidence that BCG vaccination was associated with reduced all-cause morality with some evidence that this may have been due to reduced TB mortality. I found little evidence for any other association with TB outcomes, after adjusting for confounding. In Chapter \@ref(sutherland), I explored some of the modelling evidence that was used by policy makers to assess the impact of ending school-age universal BCG vaccination. I found that the previous approach had underestimated the amount of uncertainty surrounding the effect estimates. Using newly available data, I also found that ending universal school-age BCG vaccination was projected to result in greater number of notifications in the UK born than previously thought. These findings were confirmed in Chapter \@ref(direct-eff), where I evaluated the evidence in the surveillance data that the change in policy had impacted TB incidence rates in the target populations, using Possion and negative binomial models. However, in this chapter, I found that any increase in TB notifications in the UK born was likely far outweighed by reductions in the number of notifications in the non-UK born. Using this approach, I was unable to rule out an unrelated policy change as the cause cause of this reduction. Finally, in Chapter \@ref(model-results), I forecast the impact of various vaccination scenarios using a dynamic TB model that was developed and fitted in Chapter \@ref(model-development) and Chapter \@ref(model-fitting). I found that the BCG schools' scheme was projected to reduce overall TB incidence compared to both neonatal and no vaccination over a range of time horizons. However, neonatal vaccination reduced incidence in children compared to any other scenario  although it had little impact in any other age group. No vaccination programme evaluated had an impact on incidence in older adults. The results from Chapter \@ref(model-results) are preliminary as the model on which they were based was a very poor fit to the observed data (Chapter \@ref(model-fitting)).

@@

This thesis has used multiple methods, and data sources, to explore the impact of the 2005 change in BCG policy. This multi-method approach allows for more certainty in the findings than if only a single approach had been used. A limitation of the work in this thesis is that all results were based on a single surveillance dataset. Surveillance data is subject to multiple bias issues (see Chapter \@ref(data) for details). Ideally, multiple different data types would have been used to more effectively triangulate the impact of the change in policy.  Datasets that would have added value include: notification data from other countries that also changed BCG policy, regional datasets with more reliable data BCG status and year of BCG, and data on BCG coverage and eligibility. However, the data used in this thesis represents the best available. No similarly thorough use of an equivalent data source exists for TB. Another limitation of the work in this thesis is the very poor fit of the dynamic model, developed in Chapter \@ref(model-development), to the observed data. This made drawing conclusions from it difficult. On the other hand, the model fitting presented in Chapter \@ref(model-fitting) represented one of the only examples of fitting a very complex model to data using robust bayesian, plug and play, approaches. The lack of success is itself a useful result from which much can be learned about the usability of the fitting tools and the upper limits on model complexity. A major strength of the work in this thesis is the attention that has been paid to make it both open and reproducible. Hopefully, this will allow these findings to be more easily validated, and built upon, by others. Finally the work in this thesis generated several tools as a by-product of the main research question. These tools have benefited other research projects.

@@

This thesis has highlighted the trade-off between vaccinating those at school-age and neonates in a setting where the waning of BCG effectiveness when given later in life is minimal. Whilst policy makers were previously aware of this trade-off, their was little quantitative evidence exploring it explicitly. Globally, BCG policy does not account for areas where the BCG vaccine may be equally effective regardless of when it is given.[@TheWorldHealthOrganization:2018va] Future BCG policy should consider these factors. In addition, new TB vaccines are in development that may be less susceptible to waning effectiveness when given later in life over a greater geographic area.[@Medicine2013] The findings from this thesis may be applicable to these new vaccines in areas where the BCG vaccine is currently known to be ineffective when given later in life. This may mean that these newly developed vaccines may be better targeted at those at school-age, rather than neonates, depending on the duration of protection that they provide. The work from Chapter \@ref(model-development) and Chapter \@ref(model-fitting) may be particularly suitable to adaption for this use case. This thesis has also explored the potential benefits of BCG vaccination on TB outcomes. The evidence of a reduction in all-cause mortality in TB cases may add additional weight to the argument that wider vaccination maybe more cost effective than previously thought in low incidence countries. These findings may also be used to drive vaccine uptake as they provide additional incentives for vaccination. Finally, this thesis has shown that the impact of the BCG policy has varied depending on UK birth status. This may further strengthen the case for varying vaccination policies depending on the country of origin of the target of vaccination policy, and their immediate families country of origin.

@@

Open reproducible research has been a primary focus of this thesis. A version controlled archive of this thesis is available from GitHub^[Thesis GitHub: https://github.com/seabbs/thesis], with a formatted version available on my personal site^[Thesis website: https://www.samabbott.co.uk/thesis]. This thesis relies on data from the Enhanced TB Surveillance system and the Labour Force Survey. The cleaning and munging of this data has been standardised as an R package, `tbinenglanddataclean`^[`tbinenglanddataclean`: https://www.samabbott.co.uk/tbinenglanddataclean/], and is available for download. All chapters that contain analysis are linked to their own GitHub repositories, each of which is fully reproducible (discounting the raw data which cannot be released due confidentiality reasons). Literate coding was used to link analysis code with documentation using the R tool chain. An R package, `prettypublisher`^[`prettypublisher`: https://github.com/seabbs/prettypublisher], was developed to augment these tools. Where possible open source tooling has been used to provide a working analytical environment for each chapter. All chapters that have been peer reviewed, or are undergoing peer review, have been preprinted. The model developed in Chapter \@ref(model-development) has been released as an R package along with the fitting pipeline developed in Chapter \@ref(model-fitting). Tools used to develop the figures in Chapter \@ref(background), using World Health Organization data, were expanded into an R package (see Chapter \@ref(getTBinR)). Tooling developed alongside this thesis follows open source best practices. See Chapter \@ref(introduction) for details of the open source projects developed as part of this thesis.

@@

Public engagement has been a constant theme throughout my doctoral work. This is closely linked with the previous aim of open and reproducible research. Effort has been taken so that all peer reviewed content is available for the wider public, with Twitter used to disseminate findings. Where appropriate, interactive applications have been developed that seek to explore some of the key findings of this thesis, as well as teaching more theoretical concepts used throughout (see Chapter \@ref(introduction)). Numerous case studies have also been produced that outline these theoretical concepts using some of the open source tools developed alongside this thesis^[See my personal site: https://www.samabbott.co.uk]. These tools were themselves developed to lower the barrier of entry to infectious disease research. One of these tools, `idmodelr`, has been released to CRAN. Finally, in 2017 I spent a week at the Green Man Festival exploring the mathematics of vaccination with the general public. This made use of several simple games, as well as an interactive online tool^[Available here: https://github.com/seabbs/pebblegame].

@@

The finding that BCG vaccination may reduce mortality in TB cases from Chapter \@ref(beneficial-bcg-out) require validation in other data sources and settings. A larger sample size may be required in order to unpick the association between BCG vaccination and the cause of mortality. These findings could also be included in a cost effectiveness study of the BCG vaccine. The dynamic model developed in Chapter \@ref(model-development) and fitted in \@ref(model-fitting) did not fit the observed data well. Additional compute time is needed to diagnosis whether this is a limitation of the fitting technique used or if they model itself was a poor fit for the data. Additional strategies for exploring this issue are discussed in Chapter \@ref(model-fitting). In addition this modl does not currently include targeted vaccination of neonates and is only fitted to data up to 2004. In order to be able to more accurately explore current, and future, vaccination policy the extension of the model to the present data would be required. This would potentially be of great use for policy makers. An alternative would be to develop a comparable model in different settings. This would allow the generalisability of the findings to be explored. The dynamic model could also be further generalised to include hypothetical future vaccines with differing characteristics. This would allow vaccine characteristics and optimal deployment strategies to be explored, via simulations, ahead of further development. Both `getTBinR` and `idmodelr` have active user bases and further developments are planned. This includes: additional tooling, documentation, and case studies. Further development of several of the interactive tools discussed in Chapter \@ref(introduction) is also planned.

@@

This thesis has provided new evidence regarding the use of BCG vaccination in England. A simulation study that was used as part of the quantitative evidence for the change in policy was recreated and updated. The results from this updated model suggested that the change in policy was likely to have a greater impact on the UK born, at school-age, than previously thought. This finding was supported by a regression modelling study on the impact on TB incidence rates from the policy change. However, this study also found that the change in policy was associated with some benefits in UK born neonates and a much larger reduction in TB incidence rates in both non-UK born neonates and those at school-age. An additional regression study looking at the possible link between BCG vaccination and improved TB outcomes found some evidence that BCG vaccination was associated with reduced all-cause mortality, with little evidence of any other benefits. This result strengthens the case for wider vaccination. Additionally, a dynamic model of TB transmission was developed to provide a more detailed tool for evaluating the impact of the change in policy. Although this model was a poor fit to the observed data there was some evidence that continuing school-age vaccination would have resulted in fewer overall TB cases but that neonatal vaccination reduced TB incidence in children compared to continuing school-age vaccination. There was also some evidence that none of the vaccination strategies evaluated impacted long term incidence rates in older adults. These findings suggest a stronger case for the use of the BCG vaccine in school-age populations; in areas with an equivalent level of TB transmission to England; and where the effectiveness of the BCG vaccine has been shown to not reduce with age. They also indicate that a future vaccine, without the reduced effectiveness observed in some geographic areas, may be more effectively targeted at those at school-age than at neonates. However, this depends on the potential duration of protection conferred by vaccination. In addition, they highlight the trade-off between neonatal and school-age vaccination with school-age vaccination reducing overall TB incidence but neonatal vaccination reducing incidence in young children where more severe outcomes are more common. The findings from this thesis may be of use to policy-makers to inform vaccine usage both in the UK and globally. As a by-product of the work conducted in this thesis several open source tools have been developed. These tools maybe used as learning resources, for public engangement, and as part of other research projects.
